Data organization in spreadsheets

### 2017-06-06 manuscript.Rmd
Changes from AE comments

Spreadsheets are often used as a multipurpose tool for data entry, storage,
analysis, and visualization. Most spreadsheet programs allow users to perform
all of these tasks, however we believe that spreadsheets are best suited to data
entry and storage, and that analysis and visualization should happen separately.
Analyzing and visualizing data in a separate program reduces the risk of
contaminating or destroying the raw data in the spreadsheet.

@@

@@

When formatting dates, we strongly recommend using the global
"ISO 8601" standard, `YYYY-MM-DD`, such as `2013-02-27`. (See the related xkcd
comic, <https://xkcd.com/1179>.)

But note that Microsoft Excel does bizarre things with dates
(see <https://storify.com/kara_woo/excel-date-system-fiasco>). It stores them
internally as an integer, counting the days since 1900-01-01. Wait, that is only
for Windows; on Macs, it counts the days since 1904-01-01. So, be careful to
check that the dates haven't been mangled when your data come out of Excel.

@@

However, if you do this on columns that _already_ contain dates, Excel will
convert them to a text value of their underlying integer representation (i.e.
days since 1900-01-01 or 1904-01-01).

@@

Alternatively, you could
create three separate columns
with year, month, and day. Those will be ordinary integers, and so
Excel won't mess them up.

@@

Figure 2 contains two examples of spreadsheets with some empty cells.
In Figure 2A, cells were left blank when a single value was meant to be
repeated multiple times. Please don't do this!
It is additional work for the analyst to determine the implicit values for these cells.
Moreover, if the rows are sorted at some point, the
date column will be completely mangled.

@@

## Additional considerations

Here are a few other things to avoid, not previously mentioned or
deserving further emphasis:

We once had a file with gene expression microarray data where the gene
identifiers were long integers. It had gone through Excel at some
point, and the identifier `1000000` had been changed to `1e6`, which
of course didn't match what was in the file with gene annotations.

We dislike the "Split" window in Excel (from the menu bar, Window
→ Split). A lot of our collaborators seem to use it, but it drives us
batty. To get rid of it, use Window → Remove Split. On the other
hand, we like "Freeze Panes". Go to cell B2 and click Window → Freeze
Panes. Then when you scroll down or to the right, the values in the
top row (which contain the column names) and the first column (which
often contain individual identifiers) remain shown.

The Data Carpentry lesson on using spreadsheets points to a common
mistake of not filling in zeroes (see <http://bit.ly/datacarp_scm>):
That one might enter only the non-zero values and leave the
cells-that-would-be-zero blank. *Don't do that!* Zeros are data, and
you will need them.

>>>>

Spreadsheets are often used as a multipurpose tool for data entry, storage,
analysis, and visualization. Most spreadsheet programs allow users to perform
all of these tasks, however we believe that spreadsheets are best suited to data
entry and storage, and that analysis and visualization should happen separately.
Analyzing and visualizing data in a separate program, or at least in a
separate copy of the data file, reduces the risk of
contaminating or destroying the raw data in the spreadsheet.

@@

For an existing dataset whose arrangement could be improved, we
recommend against applying tedious and potentially error-prone
hand-editing to revise the arrangement. Rather, we hope that the
reader might apply these principles when designing the layout for
future datasets.

@@

When entering dates, we strongly recommend using the global
"ISO 8601" standard, `YYYY-MM-DD`, such as `2013-02-27`. (See the related xkcd
comic, <https://xkcd.com/1179>.)

But note that Microsoft Excel does bizarre things with dates
(see <https://storify.com/kara_woo/excel-date-system-fiasco>). It stores them
internally as an number, counting the days since 1900-01-01. Wait, that is only
for Windows; on Macs, it counts the days since 1904-01-01. So, be careful to
check that the dates haven't been mangled when your data come out of Excel.

@@

However, if you do this on columns that _already_ contain dates, Excel will
convert them to a text value of their underlying numeric representation (i.e.
days since 1900-01-01 or 1904-01-01).

@@

Alternatively, you could
create three separate columns
with year, month, and day. Those will be ordinary numbers, and so
Excel won't mess them up.

@@

Figure 2 contains two examples of spreadsheets with some empty cells. In Figure
2A, cells were left blank when a single value was meant to be repeated multiple
times. Please don't do this! It is additional work for the analyst to determine
the implicit values for these cells. Moreover, if the rows are sorted at some
point there may be no way to recover the dates that belong in the empty cells.

@@
